---
title: "02 Propensity Score matching"
author: "Yuze Zheng"
date: "2026-02-26"
output: html_document
---
```{r setup, include=FALSE}
library(dplyr)
library(data.table)
library(FNN)
library(fastglm)
```

```{r}
logit_ps <- function(p) log(p / (1 - p))
keep <- "dt"
rm(list = setdiff(ls(), keep))
```

Estimate Propensity Scores using logistic regression
```{r}
ps_model <- fastglm(y = dt[,trt],
                    x = model.matrix(trt ~ L + cal_time + cal_timesqr, data = dt),
                    family = binomial(link = "logit"))
# Predicted propensity sores
dt$propensity_score <- predict(ps_model,model.matrix(trt ~ L + cal_time + cal_timesqr, data = dt), type = "response")

# View first rows
head(dt)
```

```{r}
dt
```

Matching, here we did 1 to 1 matching, with replacement
but this method takes too long so here 
so ther is the another K-nears-neib attempt use package RANN, which perform nearest neighbors for the Euclidean (L2) metric.
Here is it is a **1 to 1** matching with **replacement**
Cliper 
Here we use 
$$
\text{caliper} = 0.2 \times \operatorname{SD}\big(\operatorname{logit}(PS)\big)
$$
citation:Austin, Peter C. "Optimal caliper widths for propensity‐score matching when estimating differences in means and differences in proportions in observational studies." Pharmaceutical statistics 10.2 (2011): 150-161.
```{r}
#loop initiation 
matched_list <- list()

for (i in 1:length(unique(dt$cal_time))){
  dat <- dt %>% filter(cal_time == i-1)
  # Split treated and control groups
  treated <- dat %>% filter(trt == 1)
  control <- dat %>% filter(trt == 0)
  
  control_ps  <- as.matrix(control$propensity_score)
  treated_ps  <- as.matrix(treated$propensity_score)
  #nearest neighbor matching
  nn_result <- get.knnx(
    data = control_ps,      
    query = treated_ps,     
    k = 1
  )
  
  #set up cliper 
  sd_logit <- sd(logit_ps(c(treated$propensity_score,
                             control$propensity_score)))
  
  caliper <- 0.2 * sd_logit
  distances <- abs(treated_ps - control_ps[nn_result$nn.index])
  #cliper filter
  valid_match <- distances <= caliper
  valid_match <- as.vector(valid_match)
  # only keep calipered treated
  matched_treated <- treated[valid_match,]
  matched_control <- control[nn_result$nn.index[valid_match], ]
  # Combine matched data
  matched_data <- rbind(matched_treated,
    matched_control)
  #append 
  matched_list[[i]] <- matched_data
  }
```

```{r}
#combine list back to data.frame
ps_matched = do.call(rbind,matched_list)
ps_matched <- ps_matched[order(ps_matched$id,ps_matched$cal_time),]
head(ps_matched)
```

results check history gram
```{r}
par(mfrow=c(2,2))
p1<-dt%>% filter(trt == 0) %>% pull(propensity_score) %>% 
  hist(main = "Raw Control",xlab = "Propensity Score",col = 'white')
p3<- dt%>% filter(trt == 1) %>% pull(propensity_score) %>% 
  hist(main = "Raw Treated",xlab = "Propensity Score",col = 'white')
p2<-ps_matched %>% filter(trt == 0) %>% pull(propensity_score) %>% 
  hist(main = "Matched Control",xlab = "Propensity Score",col = 'grey')
p4 <-ps_matched %>% filter(trt == 1) %>% pull(propensity_score) %>% 
  hist(main = "Matched Treated",xlab = "Propensity Score",col = 'grey')
```
```{r}
dt%>% filter(trt == 1) %>% dim()
ps_matched %>% filter(trt == 1) %>% dim()
```

