---
title: "02 Propensity Score matching"
author: "Yuze Zheng"
date: "2026-02-26"
output: html_document
---
```{r setup, includ`e=FALSE}
library(dplyr)
library(data.table)
library(FNN)
library(fastglm)
```

```{r}
keep <- "dt"
rm(list = setdiff(ls(), keep))
logit_ps <- function(p) log(p / (1 - p))
```
# ── Flag new initiators ──────────────────────────────────────────────────
new_initiator == 1 only in the very first month a person starts treatment.
shift(trt, 1, fill=0) looks at the PREVIOUS row's trt value for each person.
If trt just flipped from 0 to 1, this is a new initiation.
```{r}
df <- dt %>%
  group_by(id) %>%
  arrange(cal_time) %>%
  mutate(
    new_initiator = as.integer(trt == 1 & lag(trt, 1, default = 0) == 0)
  ) %>%
  ungroup()
# View first rows
head(df)

# Quick check — each person should have at most 1 new_initiator row
df %>%
  group_by(id) %>%
  summarise(n_init = sum(new_initiator)) %>%
  count(n_init)
```

# ── Fit one logistic model with time included ────────────────────────────
# We model the probability of being a NEW initiator at each time point.
# Covariates: L (baseline confounder), cal_time, cal_timesqr (allows
# the initiation probability to curve over time, not just be linear).
#
# We fit on all rows where the person is still untreated — because only
# untreated people are "at risk" of initiating.
```{r}
ps_data <- df %>% filter(trt == 0 | new_initiator == 1)

ps_model <- glm(
  new_initiator ~ L + cal_time + cal_timesqr,
  data   = ps_data,
  family = binomial(link = "logit")
)
```

```{r}
#prediction
ps_data$ps <- predict(ps_model, newdata = ps_data, type = "response")
#data set merging
df <- df %>%
  left_join(ps_data %>% select(id, cal_time, ps), by = c("id", "cal_time"))
# Quick check — PS should be non-missing for all untreated and new initiator rows
summary(is.na(df %>% filter(trt == 0 | new_initiator == 1) %>% select(ps)))
```
# loop over & mathching 
Matching, here we did 1 to 1 matching, with replacement
but this method takes too long so here 
so ther is the another K-nears-neib attempt use package RANN, which perform nearest neighbors for the Euclidean (L2) metric.
Here is it is a 1 to 1 matching with replacement
Cliper 
Here we use 
$$
\text{caliper} = 0.2 \times \operatorname{SD}\big(\operatorname{logit}(PS)\big)
$$
citation:Austin, Peter C. "Optimal caliper widths for propensity‐score matching when estimating differences in means and differences in proportions in observational studies." Pharmaceutical statistics 10.2 (2011): 150-161.
```{r}
matched_list  <- list()   # stores matched pairs from each time point
pair_counter  <- 0        # gives each pair a unique number

for (t in sort(unique(df$cal_time))) {
  
  # Slice the dataset to just this time point
  dat <- df %>% filter(cal_time == t)
  
  # ── Treated pool: new initiators only ──────────────────────────────────────
  # These are the people who just started treatment this month.
  # They get an index date of t.
  treated <- dat %>% filter(new_initiator == 1)
  
  # If nobody starts treatment this month, skip to the next month
  if (nrow(treated) == 0) next
  
  # ── Control pool: all three eligibility conditions ─────────────────────────
  # Condition 1 — trt == 0:        not yet treated
  # Condition 2 — no outcome yet:  still at risk of the outcome
  # Condition 3 — not reused:      not already serving as a control
  control <- dat %>%
    filter(
      trt == 0,
      is.na(outcome) | outcome == 0
    )
  
  # If no eligible controls this month, skip
  if (nrow(control) == 0) next
  
  # ── Caliper ────────────────────────────────────────────────────────────────
  # The caliper is 0.2 × SD of the logit-PS across both groups at this
  # time point. This is the Austin (2011) recommendation.
  # We compute it fresh each month because the PS distribution changes.
  all_ps   <- c(treated$ps, control$ps)
  sd_logit <- sd(logit_ps(all_ps))
  
  # Guard: if SD is 0 or NA (can happen at very late time points with tiny
  # pools), skip this month — matching would be meaningless
  if (is.na(sd_logit) || sd_logit == 0) next
  
  caliper <- 0.2 * sd_logit
  
  # ── Nearest-neighbour matching ─────────────────────────────────────────────
  # get.knnx finds the single nearest control (k=1) for each treated person,
  # measured by Euclidean distance on the PS scale.
  treated_ps <- as.matrix(treated$ps)
  control_ps <- as.matrix(control$ps)
  
  nn_result  <- get.knnx(data = control_ps, query = treated_ps, k = 1)
  # this is a one to one matching we will try 1:2 -1:3 later 
  
  # ── Caliper filter ─────────────────────────────────────────────────────────
  # Compute distance between each treated person and their proposed match.
  # as.vector() on both sides prevents the matrix recycling bug we discussed.
  matched_ctrl_ps <- as.vector(control_ps[nn_result$nn.index, , drop = TRUE])
  distances       <- abs(as.vector(treated_ps) - matched_ctrl_ps)
  
  # Keep only pairs whose PS difference is within the caliper
  valid <- distances <= caliper
  if (sum(valid) == 0) next # skip to next month of
  
  # ── Extract valid matched pairs ────────────────────────────────────────────
  matched_treated <- treated[valid, ]
  matched_control <- control[nn_result$nn.index[valid], ]
  
  # ── Assign pair IDs ────────────────────────────────────────────────────────
  # Each matched pair gets a unique pair_id.
  # This is essential for strata() in the Cox model later.
  n_pairs      <- nrow(matched_treated)
  new_pair_ids <- paste0("pair_", (pair_counter + 1):(pair_counter + n_pairs))
  pair_counter <- pair_counter + n_pairs
  
  matched_treated$pair_id    <- new_pair_ids
  matched_control$pair_id    <- new_pair_ids
  
  # ── Record the index date for both members of each pair ───────────────────
  # Both people share the same index date — the month the treated person
  # initiated treatment. This is what they will reset their follow-up to.
  matched_treated$index_date <- t
  matched_control$index_date <- t
  
  # ── Save this month's pairs ────────────────────────────────────────────────
  matched_list[[as.character(t)]] <- rbind(matched_treated, matched_control)
}

```

results check history gram
```{r}
# ── Combine matched_list into one dataframe ───────────────────────────────
ps_matched <- bind_rows(matched_list)

# ── 2×2 PS overlap check ─────────────────────────────────────────────────
par(mfrow = c(2, 2))

df %>% filter(trt == 0) %>% pull(ps) %>%
  hist(main = "Raw Control",     xlab = "Propensity Score", col = "white")

df %>% filter(trt == 1) %>% pull(ps) %>%
  hist(main = "Raw Treated",     xlab = "Propensity Score", col = "white")

ps_matched %>% filter(trt == 0) %>% pull(ps) %>%
  hist(main = "Matched Control", xlab = "Propensity Score", col = "grey")

ps_matched %>% filter(trt == 1) %>% pull(ps) %>%
  hist(main = "Matched Treated", xlab = "Propensity Score", col = "grey")
```
```{r}
ps_matched %>% arrange(id,cal_time)
# ── 3c. Combine all months into one dataset ──────────────────────────────────

cat("Total matched pairs:", pair_counter, "\n")
cat("Total rows in matched dataset:", nrow(ps_matched), "\n")
```

