---
title: "03 Cox model New"
author: "Yuze Zheng"
date: "2026-03-27"
output: html_document
---
```{r}
load('after_new_02.RData')
```

# ── Build event time look up ──────────────────────────────────────────────
# For each person, find the first cal_time where outcome == 1.
# If they never have an outcome, event_time is NA and had_event is FALSE.
# We use this in Section 4 to build follow-up time after matching.
```{r}
event_lookup <- dt %>%
  group_by(id) %>%
  summarise(
    event_time = min(cal_time[outcome == 1], na.rm = TRUE),
    had_event  = any(outcome == 1)
  ) %>%
  mutate(event_time = ifelse(is.infinite(event_time), NA_real_, event_time))

# ── Get each person's treatment initiation time from the full dataset ────────
# trt_time = the first cal_time where new_initiator == 1
# For people who never initiate, this will be NA

trt_lookup <- df %>%
  group_by(id) %>%
  summarise(
    trt_start_time = ifelse(
      any(new_initiator == 1),
      min(cal_time[new_initiator == 1]),
      NA_real_
    )
  )
```

# ── 4a. Join event information onto the matched dataset ──────────────────────
# event_lookup tells us WHEN each person had their outcome (if at all).
# We join by id so each row in ps_matched gets that person's event info.
```{r}
ps_matched <- ps_matched %>%
  left_join(event_lookup, by = "id") %>%
  left_join(trt_lookup,   by = "id")
```

# ── 4b. Compute follow-up time and outcome flag ──────────────────────────────
```{r}
max_time <- 60 # end of study = last observed time point

ps_matched <- ps_matched %>%
  mutate(

    # For CONTROLS: censor at whichever comes first —
    #   their treatment initiation, their outcome, or study end
    # For TREATED:  their protocol was "initiate at index date" — 
    #   no further censoring needed, follow until outcome or study end

    pp_censor_time = ifelse(
      trt == 0,
      trt_start_time,   # controls: censor when they start treatment
      NA_real_          # treated:  no per-protocol censoring
    ),

    # end_time = earliest of: outcome, per-protocol censor, study end
    # pmin() with na.rm=TRUE ignores NA (e.g. treated people have no
    # pp_censor_time, never-treated controls have no trt_start_time)
    end_time = pmin(
      ifelse(had_event, event_time, max_time),
      ifelse(!is.na(pp_censor_time), pp_censor_time, max_time),
      na.rm = TRUE
    ),

    # Reset clock to zero from index date
    follow_up = end_time - index_date,

    # Event only counts if it happened AFTER index date
    # AND before or at the per-protocol censoring time
    outcome_flag = as.integer(
      had_event &
      event_time > index_date &
      (trt == 1 | is.na(trt_start_time) | event_time <= trt_start_time)
    )
  )
```

# ── 4c. Quick sanity checks ──────────────────────────────────────────────────
```{r}
# follow_up should all be >= 0
stopifnot(all(ps_matched$follow_up >= 0))

# Every pair_id should appear exactly twice (one treated, one control)
pair_counts <- ps_matched %>% count(pair_id)
stopifnot(all(pair_counts$n == 2))

# Check event rates by group
ps_matched %>%
  group_by(trt) %>%
  summarise(
    n          = n(),
    n_events   = sum(outcome_flag),
    event_rate = mean(outcome_flag)
  )
```

```{r}
# Check how many controls were per-protocol censored
ps_matched %>%
  filter(trt == 0) %>%
  summarise(
    n_total          = n(),
    n_pp_censored    = sum(!is.na(trt_start_time) &
                           trt_start_time > index_date &
                           trt_start_time <= max_time),
    pct_pp_censored  = mean(!is.na(trt_start_time) &
                            trt_start_time > index_date &
                            trt_start_time <= max_time) * 100
  )
```

# ── 5a. Fit the stratified Cox model ─────────────────────────────────────────
this is more itt approach
Surv(follow_up, outcome_flag) — the outcome variable
trt                           — the exposure (1=treated, 0=control)
strata(pair_id)               — compares within pairs, not across them
cluster(id)                   — corrects SEs for reused controls
```{r}
library(survival)
cox_model <- coxph(
  Surv(follow_up, outcome_flag) ~ trt + strata(pair_id) + cluster(id),
  data   = ps_matched,
  ties   = "efron"    # Efron approximation handles tied event times well
)

summary(cox_model)
```
# ── 5b. Extract the hazard ratio and confidence interval ─────────────────────
#* exp(coef) gives the hazard ratio.
#* A HR < 1 means treatment reduces the hazard of the outcome.
#* A HR > 1 means treatment increases it.
```{r}
hr_table <- broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value)

print(hr_table)
```
# ── 5c. Plot Kaplan-Meier survival curves by treatment group ─────────────────
#* This gives a visual check — do the curves separate in the expected direction?
```{r}
km_fit <- survfit(
  Surv(follow_up, outcome_flag) ~ trt,
  data = ps_matched
)

plot(
  km_fit,
  col     = c("steelblue", "tomato"),
  lwd     = 2,
  xlab    = "Months since index date",
  ylab    = "Probability of remaining event-free",
  main    = "Kaplan-Meier curves after PSM"
)
legend("bottomleft",
       legend = c("Control", "Treated"),
       col    = c("steelblue", "tomato"),
       lwd    = 2)
```